Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression

### 2018-01-16 paper/paper.tex
Change notation for spins: S_i -> \chi_i

With these relations the waveform only depends on 5 intrinsic parameters $\bx=\{q, S_A, S_B, \Lambda_2^A,\Lambda_2^B\}$. We note that this list does not include the total mass $M$ of the system. The point-mass part of the model is scale-invariant, and the tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$ and $\Lambda_2^{A,B}$. Finally, the simple model of the post-merger signal that we employ rescales with $M$ as well. It is not clear at what point in the transition from the inspiral to the post-merger this approximation breaks down. For example, whether the merging binary undergoes prompt collapse or forms a hypermassive remnant depends sensitively on the total mass. A set of NR simulations with a fixed mass ratio and tidal parameters but different total masses (and therefore different EOSs in order to keep the tidal parameters fixed) will be needed to determine the time when this approximation breaks down.

@@

For aligned-spin waveforms, we can solve this problem by using the fact that the waveform can be approximated with the analytic \texttt{TaylorF2} waveform $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$. This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2
\begin{align}
\Delta\ln(A)(Mf; \bx) &= \ln\left(\frac{ A(Mf; \bx) }{ A_{\rm F2}(Mf; \bx)}\right), \label{eq:ampres}\\
\Delta\Phi(Mf; \bx) & = \Phi(Mf; \bx) - \Phi_{\rm F2}(Mf; \bx), \label{eq:phaseres}
\end{align}
such that
\begin{align}
\tilde h(Mf; \bx) = \tilde h_{\rm F2}(Mf; \bx) e^{\Delta\ln(A)(Mf; \bx) + i  \Delta\Phi(Mf; \bx)}.
\label{eq:hdecomp}
\end{align}
These residuals are also shown in Fig.~\ref{fig:hoff}. For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that interpolation errors will not lead to a negative amplitude for the reconstructed waveform (Eq.~\eqref{eq:hdecomp}). In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency, the log-ratio captures this behavior. Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies. In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $S_1$. \MP{TODO: replace $S_i$ by $\chi_i = \hat L_N \cdot \vec S_i / m_i^2$.} These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work. The exception is the amplitude residual at very high frequencies where the amplitude is small and numerical noise is significant.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{dhofs.pdf}
\caption{Cross sections of the amplitude and phase residuals as functions of $S_1$ at three nodes $MF_j$. The waveforms are generated for 51 values of $S_1$ shown as dots with the other parameters held constant at $\{q, S_2, \Lambda_1, \Lambda_2\} = \{0.6, 0.2, 2000, 1000\}$. Also shown are 1-dimensional cross sections of the interpolated residuals using Gaussian process regression (solid black) as well as their estimated $1\sigma$ uncertainty (shaded gray region).}
\label{fig:dhofs}
\end{figure}

The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
\begin{equation}
A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} x^{-7/4} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)x \right],
\end{equation}
where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{LAL}~\cite{lal}. The phase has the schematic form
\begin{align}
\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, S_1, S_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
\end{align}
We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order~\cite{BoheMarsatBlanchet2013}, and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order~\cite{VinesFlanaganHinderer2011}. These terms for the phase are exactly as used in the \texttt{LAL} waveform \texttt{TaylorF2}. \MP{add git hash? LAL's TF2 will change in the future when new terms become available.}

@@

To validate that the waveforms are sufficiently conditioned, we plot cross-sections of the residuals in Fig.~\ref{fig:dhofs} for fixed frequencies as functions of one of the waveform parameters $S_1$. The fact that the residuals are smooth functions of the waveform parameters indicates that most numerical noise has been removed. The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.052; \bx)$, where the amplitude is very small and dominated by numerical noise. 
\MP{The discussion of Fig.~\ref{fig:dhofs} appears twice in the text; here and one section earlier.}

@@

Our final kernel takes the form
\begin{equation}
k(\bx_i, \bx_j) = \sigma_f^2 k_{\rm Matern}^{\nu=5/2}(r) + \sigma_n^2 \delta_{ij}
\end{equation}
where $\sigma_f$ is a scale factor that describes the range of values that $g(\bx)$ takes over the domain, and $\sigma_n$ is a noise parameter. The white noise kernel $\sigma_n^2 \delta_{ij}$ (also called a nugget) is $\sigma_n^2$ when $\bx_i = \bx_j$ and zero otherwise, and parameterizes noise in the data $\by$. In our case, the training set waveforms have numerical noise that we will estimate by optimizing the hyperparameters. The full set of hyperparameters is now  $\btheta = \{\sigma_f, \ell_q, \ell_{S_1}, \ell_{S_2}, \ell_{\Lambda_1}, \ell_{\Lambda_2}, \sigma_n\}$.

@@

We build an initial training set with 128 waveforms sampled with an LHD in our 5-dimensional parameter space given by $q\in[1/3, 1]$, $S_1, S_2 \in [-0.5, 0.5]$ and $\Lambda_1, \Lambda_2 \in [0, 5000]$. In addition, we find empirically that the GPR uncertainty estimates (Eq.~\eqref{eq:var} and Eq.~\eqref{eq:errorrms} below) are largest at the corners of the parameter space, so we also sample the 32 corners. We construct our initial surrogate with these 160 waveforms shown in Fig.~\ref{fig:LHD}.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{trainingset3d.pdf}\\
\includegraphics[width=0.49\textwidth]{trainingset2d.pdf}
\caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, S_1, S_2\}$ (top) and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom). The 160 blue circles were used for the initial training set (32 corners + 128 LHD points). The 200 red squares were generated using uncertainty sampling from the GPR error estimate (Eq.~\eqref{eq:errorrms}). The final 1000 green triangles were drawn uniformly from the parameter space.
}
\label{fig:LHD}
\end{figure}

>>>>

With these relations the waveform only depends on 5 intrinsic parameters $\bx=\{q, \chi_A, \chi_B, \Lambda_2^A,\Lambda_2^B\}$. The aligned component spins are defined as $\chi_i = \hat L_N \cdot \vec S_i / m_i^2$ where $\vec S_i$ are the dimensionfull spin vectors, $\hat L_N$ the Newtonian orbital angular momentum and $m_i$ the component masses.
We note that this list does not include the total mass $M$ of the system. The point-mass part of the model is scale-invariant, and the tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$ and $\Lambda_2^{A,B}$. Finally, the simple model of the post-merger signal that we employ rescales with $M$ as well. It is not clear at what point in the transition from the inspiral to the post-merger this approximation breaks down. For example, whether the merging binary undergoes prompt collapse or forms a hypermassive remnant depends sensitively on the total mass. A set of NR simulations with a fixed mass ratio and tidal parameters but different total masses (and therefore different EOSs in order to keep the tidal parameters fixed) will be needed to determine the time when this approximation breaks down.

@@

For aligned-spin waveforms, we can solve this problem by using the fact that the waveform can be approximated with the analytic \texttt{TaylorF2} waveform $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$. This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2
\begin{align}
\Delta\ln(A)(Mf; \bx) &= \ln\left(\frac{ A(Mf; \bx) }{ A_{\rm F2}(Mf; \bx)}\right), \label{eq:ampres}\\
\Delta\Phi(Mf; \bx) & = \Phi(Mf; \bx) - \Phi_{\rm F2}(Mf; \bx), \label{eq:phaseres}
\end{align}
such that
\begin{align}
\tilde h(Mf; \bx) = \tilde h_{\rm F2}(Mf; \bx) e^{\Delta\ln(A)(Mf; \bx) + i  \Delta\Phi(Mf; \bx)}.
\label{eq:hdecomp}
\end{align}
These residuals are also shown in Fig.~\ref{fig:hoff}. For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that interpolation errors will not lead to a negative amplitude for the reconstructed waveform (Eq.~\eqref{eq:hdecomp}). In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency, the log-ratio captures this behavior. Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies. In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $\chi_1$. These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work. The exception is the amplitude residual at very high frequencies where the amplitude is small and numerical noise is significant.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{dhofs.pdf}
\caption{Cross sections of the amplitude and phase residuals as functions of $\chi_1$ at three nodes $MF_j$. The waveforms are generated for 51 values of $\chi_1$ shown as dots with the other parameters held constant at $\{q, S_2, \Lambda_1, \Lambda_2\} = \{0.6, 0.2, 2000, 1000\}$. Also shown are 1-dimensional cross sections of the interpolated residuals using Gaussian process regression (solid black) as well as their estimated $1\sigma$ uncertainty (shaded gray region).}
\label{fig:dhofs}
\end{figure}

The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
\begin{equation}
A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} x^{-7/4} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)x \right],
\end{equation}
where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{LAL}~\cite{lal}. The phase has the schematic form
\begin{align}
\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, \chi_1, \chi_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
\end{align}
We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, the aligned spin terms ${\rm Spin}(\eta, \chi_1, \chi_2)$ to 3PN order~\cite{BoheMarsatBlanchet2013}, and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order~\cite{VinesFlanaganHinderer2011}. These terms for the phase are exactly as used in the \texttt{LAL} waveform \texttt{TaylorF2}. \MP{add git hash? LAL's TF2 will change in the future when new terms become available.}

@@

To validate that the waveforms are sufficiently conditioned, we plot cross-sections of the residuals in Fig.~\ref{fig:dhofs} for fixed frequencies as functions of one of the waveform parameters $\chi_1$. The fact that the residuals are smooth functions of the waveform parameters indicates that most numerical noise has been removed. The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.052; \bx)$, where the amplitude is very small and dominated by numerical noise. 
\MP{The discussion of Fig.~\ref{fig:dhofs} appears twice in the text; here and one section earlier.}

@@

Our final kernel takes the form
\begin{equation}
k(\bx_i, \bx_j) = \sigma_f^2 k_{\rm Matern}^{\nu=5/2}(r) + \sigma_n^2 \delta_{ij}
\end{equation}
where $\sigma_f$ is a scale factor that describes the range of values that $g(\bx)$ takes over the domain, and $\sigma_n$ is a noise parameter. The white noise kernel $\sigma_n^2 \delta_{ij}$ (also called a nugget) is $\sigma_n^2$ when $\bx_i = \bx_j$ and zero otherwise, and parameterizes noise in the data $\by$. In our case, the training set waveforms have numerical noise that we will estimate by optimizing the hyperparameters. The full set of hyperparameters is now  $\btheta = \{\sigma_f, \ell_q, \ell_{\chi_1}, \ell_{\chi_2}, \ell_{\Lambda_1}, \ell_{\Lambda_2}, \sigma_n\}$.

@@

We build an initial training set with 128 waveforms sampled with an LHD in our 5-dimensional parameter space given by $q\in[1/3, 1]$, $\chi_1, \chi_2 \in [-0.5, 0.5]$ and $\Lambda_1, \Lambda_2 \in [0, 5000]$. In addition, we find empirically that the GPR uncertainty estimates (Eq.~\eqref{eq:var} and Eq.~\eqref{eq:errorrms} below) are largest at the corners of the parameter space, so we also sample the 32 corners. We construct our initial surrogate with these 160 waveforms shown in Fig.~\ref{fig:LHD}.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{trainingset3d.pdf}\\
\includegraphics[width=0.49\textwidth]{trainingset2d.pdf}
\caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, \chi_1, \chi_2\}$ (top) and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom). The 160 blue circles were used for the initial training set (32 corners + 128 LHD points). The 200 red squares were generated using uncertainty sampling from the GPR error estimate (Eq.~\eqref{eq:errorrms}). The final 1000 green triangles were drawn uniformly from the parameter space.
}
\label{fig:LHD}
\end{figure}
